A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-05-06 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
First round of revisions after invitation to revise and resubmit.

  \begin{abstract}
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
  systems and the economy. Public health interventions have slowed the
  spread of the virus, albeit at a high cost. How and when to ease
  restrictions to movement hinges in part on whether SARS-CoV2 will
  display seasonality due to variations in temperature, humidity, and
  hours of sunshine. Here, we address this by means of a spatio-temporal
  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
  Unrelated Regressions (SUR) allows us to model the incidence of reported
  cases of the disease per 100,000 population as an interregional
  contagion process, in addition to a function of temperature, humidity,
  and sunshine. In the analysis we also control for GDP per capita,
  percentage of older adults in the population, population density, and
  presence of mass transit systems. The results support the hypothesis
  that incidence of the disease is lower at higher temperatures and higher
  levels of humidity. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.\\
  \end{abstract}

@@

From a small outbreak linked to a live animal market at the end of 2019
to a global pandemic in a matter of weeks, the SARS-CoV2 virus has
threatened to overrun health systems the world over. In efforts to
contain the spread, numerous governments in many nations and regions
have either recommended or mandated social distancing measures, and as
of this writing, millions of people in five continents shelter in place.
There are encouraging signs that these measures have arrested the spread
of the virus where they have been implemented, and have thus helped to
keep a bad situation from becoming even worse (e.g., 2020). However,
this has come at a high cost, and the consequences for all spheres of
the economy, social, and cultural life have been dire (e.g., Fernandes,
2020; Luo and Tsang, 2020). As a result, there is a sense of urgency to
anticipate the progression of the pandemic, in order to plan for
progressive lifting of restrictions to movement and social contact
(e.g., Kissler et al., 2020). Needless to say, this has become a
delicate, and politically charged, balancing act between public health
and the economy (Gong et al., 2020).

A salient question in the debate on how and when to ease social
distancing measures is whether the virus will display seasonal
variations. Earlier, diverse studies have shown the effect of
temperature and humidity on the incidence of influenza (e.g., Mäkainen
et al., 2009; Jaakkola et al., 2014; Kudo et al., 2019). Jaakkola et
al.~(2014), for example, found that a decrease of temperature and
absolute humidity precedes the onset of symptoms of influenza A and B
viruses by 3 days in places where the temperature is low. After the
2002-2004 outbreak of SARS, researchers also began to investigate the
relationship between these factors and SARS-CoV. In this way, Casanova
et al.~(2010) used two surrogates, namely the gastroenteritis (TGEV) and
mouse hepatitis viruses (MHV), to find that virus inactivation was more
rapid at temperatures of 20C than 4C, and at 40C than 20C; in terms of
humidity, these researchers reported that survival of the virus was
lower at moderate relative humidity levels. In a similar vein, but
working directly with SARS-CoV, Chan et al.~(2011) found that viability
of the virus was lost at temperatures higher than 38C and relative
humidity superior to 95\%.

While existing research on similar pathogens suggests that SARS-CoV is
more stable and potentially easier to transmit in conditions of low
temperature and low humidity, it is far from certain that this will also
be the case with the novel SARS-CoV2. If such is the case, there is the
possibility of easing restrictions to social contact as the weather
warms; however, weeks or possibly months of costly measures could become
undone if not, and the restrictions are lifted prematurely. Not
surprisingly, given the stakes involved, this issue has already
triggered a lively debate.

Some of what is thought about the possible seasonality of COVID-19 is
based on analogies to the patterns of other known respiratory viruses.
However, de Ángel Solá et al.~(2020) note that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns''
(section 4). This urges caution when extrapolating from known viruses,
and indeed, the evidence in this respect is inconclusive. At a global
scale, de Ángel Solá et al.~(2020) see less risk in the Caribean Basin;
however, Coelho et al.~(Coelho et al., 2020) warn that at least during
the exponential phase, expansion of the virus is not driven by climate.
Similarly, whereas Araujo and Naimi (2020) argue that spread of
SARS-CoV2 will likely be constrained by climate, Harbert et al.~(2020)
remain unconvinced that spatial modelling can currently discriminate the
distribution of the disease on the basis of climate, at least in the
United States. Yao et al.~(2020), examined data from China and came to
the conclusion that neither temperature nor ultraviolet indices had an
association with transmission of COVID-19. This is despite previous
research that has linked less exposure to UVB radiation to higher
prevalence and severity of acute respiratory tract infections
(Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen et
al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020). To further complicate matters,
much of the relevant work has yet to be peer-reviewed and therefore is
open to change (see for example the challenge of Harbert et al. (2020)
to Araujo and Naimi (2020)). In the United States, the National Academy
of Sciences, Engineering, and Medicine was engaged by the Office of the
Executive for guidance on this matter (see National Academies of
Sciences, Engineering and Medicine, 2020). Their conclusion summarizes
the situation well (see p.~6): ``Some limited data support a potential
waning of cases in warmer and more humid seasons, yet none are without
major limitations\ldots{}Additional studies as the SARS-CoV-2 pandemic
unfolds could shed more light on the effects of climate on
transmission.''

With the above considerations in mind, our objective with this paper is
to contribute to the knowledge basis regarding the spread of COVID-19
and the influence of environmental factors, particularly temperature,
humidity, and sunshine. We adopt a population health approach, and
report results from a spatial model of the incidence of COVID-19 in
fifty provinces in Spain, one of the countries hardest hit by the
pandemic. We combine data on reported cases of the disease with
metereological information, to create a spatio-temporal dataset covering
a period of 30 days. We then join this dataset with provincial-level
economic and demographic information to act as controls to our key
environmental variables. These data are analyzed using a spatial
Seemingly Unrelated Regressions (SUR) approach, which allows us to model
incidence of COVID-19 as a spatial contagion process.

@@

Please note that this paper is prepared as a reproducible research
document. The source R Markdown document, as well as all data and code
needed to reproduce/review/extend the analysis can be obtained from the
following anonymous Drive folder:

https://drive.google.com/open?id=1d\_40N\_QXo2Fl14r3T3CHs84IED-B1rV\_

@@

The first reported case of COVID-19 in Spain was on January 31th, when a
German tourist in the Canary Islands tested positive for the virus.
However, it was still a few weeks before the first domestic case was
reported, on February 27th in Sevilla province (Andalusia). In a short
period of time, after this relatively slow start, the outbreak flared.
By March 11th the World Health Organization (WHO) declared COVID-19
officially a pandemic. This declaration marked a turning point for the
public in Spain too. As of March 13th, the number of cases of COVID-19
reported in Spain was 4,473, with a majority of cases (1,990)
concentrated in Madrid: these numbers were at the time the worst
outbreak in Europe, after Italy. In response to the situation, on March
13th the Spanish National Government declared a state of emergency, to
go into effect on Saturday March 14th. As part of the state of emergency
restrictions to most activities were imposed, with the exception of
essential services (e.g.~food, health) and some economic subsectors of
industry and construction. A few days later, on March 17th, Spain closed
its lands borders to allow entry only to returnee nationals and
permanent residents. The lockdown was further hardened between March
30th and April 12th (including the Easter weekend of April 10th-12th)
and during this period only essential activities were allowed. During
this period, there was a dramatic reduction in overall mobility, both
within provinces as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

\hypertarget{variables}{%
\subsection{Selection of Variables}\label{variables}}

The global emergence of infectious diseases is mostly driven by
environmental, ecological, and socio-economic factors (Jones et al.,
2008). In the case of SARS-CoV2, the ecological factors include the
interaction between humans and wildlife. Once transmission of a disease
begins to happen between humans, socio-economic and environmental
factors become increasingly important. As noted in the introduction, the
focus of the paper is on environmental variables, namely temperature,
humidity, and sunshine. These variables have been implicated in the
viability and ease of transmission of similar viruses. In addition to
these variables, we also aim to use a set of controls, in the form of
specific socio-economic and demographic characteristics of each
province.

The percentaje of older adults (over 65) is the second variable that we
consider as a control. Early evidence regarding COVID-19 suggests that
the case rate mortality is higher at older ages (e.g.~The Novel
Coronavirus Pneumonia Emergency Response Epidemiology Team, 2020).
However, it is not clear that a relatively large population of older
adults necessarily translates into higher transmission of the infection.
The tool of choice in containing the spread of the disease has been
social distancing. In this respect, the evidence from the field of
transportation is that older adults tend to travel less frequently, for
shorter distances, and have higher rates of immobility (e.g., Roorda et
al., 2010; Morency et al., 2011; Sikder and Pinjari, 2012). In other
words, many older adults are, whether by preference or otherwise,
already in a form of social isolation. Social distancing during the
pandemic may actually reinforce that condition for them, as suggested by
the analysis of age-structured social contact in India, China, and Italy
of Sing and Adhikari (2020). Accordingly, our expectation is that
provinces with higher percentages of older adults will tend to have
similarly low levels of social contact, particularly since the
age-structured matrix of social contact in Spain is similar to Italy
(Prem et al., 2017).

The third population variable the we consider is population density,
since it directly affects the contact patterns and contact rates between
individuals in a population (Hu et al., 2013). The evidence available
suggests a positive relationship between the transmission of COVID-19
and population density (e.g.~cumulative incidence in urban areas like
NYC). For this reason, we anticipate a positive relationship between
population density and the incidence of the disease.

The last control variable is the presence of mass transit systems in a
province. Every province in Spain offers some form of public
transportation, however only five provinces have higher order systems of
mass mobility (e.g.~metro or subway), namely Barcelona, Madrid, Sevilla,
Valencia, and Bizkaia. Public transportation has been hypothesized to
relate to the spread of contagious disease by some researchers using
agent-based approaches and simulation (e.g., Perez and Dragicevic, 2009;
Wang et al., 2010), and while we find scant evidence of a link in the
literature, the idea is intuitively appealing. After all, unlike the
isolation that a car offers to travellers, most mass transit system are
cauldrons of social contact.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTIII in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. For our purposes, we consider positive cases reported, but
excluding symptomatic cases diagnosed by a doctor without a Polymerase
Chain Reaction (PCR) test. The Spanish National Government publishes
periodic updates at the regional level (NUTII) and the information is
also released at the provincial level as part of a collaborative project
by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
and ESRI España. This information is compiled from several sources,
mainly the official web pages of the Spanish regional goverments, as
documented in Centro de Datos
Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
In addition, we consider two sets of explanatory variables. The first
one, and the focus of this research, is set of two environmental
variables, namely temperature and humidity, which are collected from
official sources (i.g., \emph{AEMET}, the state meteorology agency, and
\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
second set provides some relevant controls for multivariate analysis,
and refers to economic and demographic attributes of the province (also
collected from official sources, i.e., INE, the national statistics
institute). Table \ref{tab:descriptive-statistics} shows the descriptive
statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset
begins on March 13, which is the first date when every province had
reported at least one case of COVID-19, and continues until April 11,
for a period of 30 days. The unit of analysis is the province. Provinces
are the equivalent of states, and are embedded in Autonomous
Communities. As an example, Cataluña is an Autonomous Community and
consists of four provinces, namely Barcelona, Gerona, Lerida, and
Tarragona. The size of the provinces is relatively large, as seen in
Table \ref{tab:descriptive-statistics}. The smallest province is
\(1,978.12km^2\) (this is smaller than Rhode Island in the US) and the
largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
in the US). While this is a relatively large degree of spatial
aggregation, reporting on COVID-19 is inconsistent at smaller
geographies, or cases are not reported at that level at all. The
analysis must therefore be considered ecological.

@@

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics} Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & ProvidencialData19\\
Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4805.98 & INE\\
Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\addlinespace
Mean Temperature & Mean temperature in province by date, in C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

@@

The Seemingly Unrelated Regression equations model (SUR hereafter) is a
multivariate econometric model used in different fields when the
structure of the data consisits of cross-sections for different time
periods. The basis of this approach is well-known since the initial
works of Zellner (1962), and has become a popular methodology included
in several econometrics textbook (e.g., Greene, 2003). To our knowledge,
Anselin (1988) was the first author to discuss SUR from a spatial
perspective. In his landmark text, Anselin discussed a model made of
``an equation for each time period, which is estimated for a cross
section of spatial units'' (p.~141). From this milestone, a large body
of research has developed to extend the classical SUR into a spatial
framework (e.g., Rey and Montouri, 1999; Lauridsen et al., 2010; Le
Gallo and Dall'Erba, 2006; López et al., 2017).

@@

A key feature of the SUR model is the dependence structure among the
vectors of residuals, namely:

@@

\hypertarget{analysis}{%
\section{Analysis}\label{analysis}}

@@

Figure \ref{fig:descriptives-temperature} shows the distribution of the
environmental variables in Spain. For ease of visualization we aggregate
the provinces by Autonomous Community. Each box-and-whisker in the
figure represents the distribution of the variable for a community over
the 30-day period. In the plot, the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. As seen in the figure, there is a
relatively wide range of values both within and between provinces over
the 30-day period examined. The top panel of the figure shows the
distribution of mean temperatures. The lowest mean temperature for a
community on any given day was 2.8C, and the highest 22.4C, for a range
of approximately 20 degrees. Likewise, there is a great deal of
variability in humidity, as seen in the middle panel of the figure,
where the lowest mean humidity for any community is 48.3 and the highest
is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
in the community. This variable displays the most variability within
communities over time, but remains relatively constant across
communities. It is important to note that these are summaries by
community, and the actual values of these variables for the provinces
display somewhat more variability.

@@

It can be seen in the figure that temperature (in its three forms) has
the highest simple correlation with incidence of COVID-19. After
temperature, GDP per capita has the highest positive correlation with
the dependent variable. The distribution of these correlations is also
quite tight over the 30-day period of the study. Hours of sunshine tends
to have a moderately high correlation with incidence of COVID-19,
however the distribution of these correlations is more spread, and in
some cases strays into negative values. A similar thing happens with
humidity, which also tends to display more day to day variation in the
correlation with the dependent variable. The percentage of older adults
shows a relatively tight distribution of day-to-day correlations, and is
negative. Population density, in contrast, tends to be negative, but is
relatively spread, and on some days, the simple correlation between
density and incidence of COVID-19 is weakly positive.

@@

\hypertarget{results-and-discussion}{%
\section{Results and Discussion}\label{results-and-discussion}}

@@

Two variables had temporally constrained coefficients. The estimated
effect of GDP per capita is to increase the incidence of COVID-19 by
0.675\% for each percentual increase of this variable (in €1,000s). In
our view, this is a measure of inertia, as provinces with higher GDP per
capita where among the first to see exponential growth in the pandemia.
Percentage of older adults has a negative effect, and each percentual
increase in this variable is associated with a relatively small
reduction of the incidence of less than 1\%.

>>>>

  \begin{abstract}
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health
  systems and the economy. Public health interventions have slowed the
  spread of the virus, albeit at a high cost. How and when to ease
  restrictions to movement hinges in part on whether SARS-CoV2 will
  display seasonality due to variations in temperature, humidity, and
  hours of sunshine. Here, we address this by means of a spatio-temporal
  analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly
  Unrelated Regressions (SUR) allows us to model the incidence of reported
  cases of the disease per 100,000 population as an interregional
  contagion process, in addition to a function of temperature, humidity,
  and sunshine. In the analysis we also control for GDP per capita,
  percentage of older adults in the population, population density, and
  presence of mass transit systems. The results support the hypothesis
  that incidence of the disease is lower at higher temperatures and higher
  levels of humidity. Sunshine, in contrast, displays a positive
  association with incidence of the disease. Our control variables also
  yield interesting insights. Higher incidence is associated with higher
  GDP per capita and presence of mass transit systems in the province; in
  contrast, population density and percentage of older adults display
  negative associations with incidence of COVID-19.
  \end{abstract}

@@

From a small outbreak linked to a live animal market at the end of 2019
to a global pandemic in a matter of weeks, the SARS-CoV2 virus (and
COVID-19, the disease it causes), has threatened to overrun health
systems around the world. In efforts to contain the spread of the
disease, numerous governments in many regions and nations have either
recommended or mandated social distancing measures, and as of this
writing, millions of people in five continents shelter in place. There
are encouraging signs that these measures have mitigated the spread of
the virus (e.g., Lancastle, 2020; Lewnard and Lo, 2020; Wilder-Smith and
Freedman, 2020). Even so, this has come at a high cost, and the
consequences for all spheres of economic, social, and cultural life have
been dire (e.g., Fernandes, 2020; Luo and Tsang, 2020). As a result,
there is a sense of urgency to anticipate the progression of the
pandemic, in order to plan for progressive lifting of restrictions to
movement and social contact (e.g., Kissler et al., 2020). Needless to
say, this has become a delicate, and politically charged, balancing act
between public health and the economy (Gong et al., 2020).

A salient question in the debate on how and when to ease social
distancing measures is whether the virus will display seasonal
variations. Existing research on similar pathogens suggests that the
virus could be more stable and potentially easier to transmit in
conditions of low temperature and low humidity. While encouraging this
is encouraging, it is important to keep in mind that ``not all seasonal
respiratory viruses experience the same spatiotemporal patterns'' (Ángel
Solá et al., 2020, sec. 4). This urges caution when extrapolating from
known viruses. The evidence in this respect is as yet inconclusive, and
although easing restrictions as the weather warms may appear tempting,
doing so prematurely could well undo weeks or possibly months of costly
measures.

It is not surprising, given the stakes involved, that this issue has
already triggered a lively debate. The current state of knowledge is was
well-summarized by the National Academy of Sciences, Engineering, and
Medicine in the U.S. (see National Academies of Sciences, Engineering
and Medicine, 2020). Engaged by the Office of the Executive for guidance
on this matter, this organization concludes that: ``{[}some{]} limited
data support a potential waning of cases in warmer and more humid
seasons, yet none are without major limitations\ldots Additional studies
as the SARS-CoV2 pandemic unfolds could shed more light on the effects
of climate on transmission'' (p.~6). To further complicate matters, much
of the relevant work has yet to be peer-reviewed (see for instance the
challenge of Harbert et al. (2020) to Araujo and Naimi (2020)).

With the above considerations in mind, our objective with this paper is
to investigate the influence of environmental factors, concretely
temperature, humidity, and sunshine, on the progression of the pandemic.
We adopt a population health approach, and report results from a
spatio-temporal model of the incidence of COVID-19 in fifty provinces in
Spain, one of the countries hardest hit by the pandemic. We combine data
on reported cases of the disease with metereological information, to
create a spatio-temporal dataset covering a period of 30 days. We then
join this dataset with provincial-level economic and demographic
information to act as controls to our key environmental variables. These
data are analyzed using a spatial Seemingly Unrelated Regressions (SUR)
approach, which allows us to model incidence of COVID-19 as a contagion
process.

@@

Please note that this paper is prepared as a reproducible research
document. The source \texttt{R} markdown document, as well as all data
and code needed to reproduce/review/extend the analysis can be obtained
from a public
repository\footnote{https://github.com/paezha/covid19-environmental-correlates}.

\hypertarget{background}{%
\section{Background}\label{background}}

The global emergence of infectious diseases is mostly driven by
environmental, ecological, and socio-economic factors (Jones et al.,
2008). In the case of SARS-CoV2, the ecological factors include the
interaction between humans and wildlife. Once transmission of a disease
begins to happen between humans, socio-economic and environmental
factors become increasingly important. As noted in the introduction, the
focus of the paper is on environmental variables, namely temperature,
humidity, and sunshine.

Much of what is known about the potential seasonality of SARS-CoV2 is a
result of research on other pathogens. Earlier, diverse studies have
shown the effect of temperature and humidity on the incidence of
influenza (e.g., Mäkainen et al., 2009; Jaakkola et al., 2014; Kudo et
al., 2019). Jaakkola et al.~(2014), for example, found that a decrease
of temperature and absolute humidity precedes the onset of symptoms of
influenza A and B viruses by 3 days in places where the temperature is
low. After the 2002-2004 outbreak of SARS, researchers also began to
investigate the relationship between these factors and SARS-CoV2
(Casanova et al., 2010; Chan et al., 2011). Casanova et al.~(2010), for
instance, used two surrogates (i.e., the gastroenteritis TGEV and mouse
hepatitis MHV viruses), to find that virus inactivation was more rapid
at temperatures of 20°C than 4°C, and at 40°C than 20°C; in terms of
humidity, these researchers reported that survival of the virus was
lower at moderate relative humidity levels. Chan et al.~(2011) also
found that viability of the virus that causes SARS is also lost at
higher temperatures (\textgreater38°C) and relative humidity superior to
95\%.

Whether these results from laboratory experiments will hold in general
remains uncertain. At a global scale, de Ángel Solá et al.~(2020) see
less risk in the Caribean Basin; however, Coelho et al.~(Coelho et al.,
2020) warn that at least during the exponential phase, expansion of the
virus is not driven by climate. Similarly, whereas Araujo and Naimi
(2020) argue that spread of SARS-CoV2 will likely be constrained by
climate, Harbert et al.~(2020) remain unconvinced that spatial modelling
can currently discriminate the distribution of the disease on the basis
of climate, at least in the United States. Yao et al.~(2020), examined
data from China and came to the conclusion that neither temperature nor
ultraviolet indices had an association with transmission of COVID-19.
This is despite previous research that has linked less exposure to UVB
radiation to higher prevalence and severity of acute respiratory tract
infections (Zittermann et al.~2016; Dąbrowska-Leonik et al.~2018; Dinlen
et al.~2016; Mathyssen et al.~2017; Esposito and Lelii 2015; Jat 2017;
Moriyama, Hugentobler, and Iwasaki 2020).

In addition to the environmental variables above, we also consider a set
of potential socio-economic and demographic confounders.

Secondly, we consider percentage of older adults (over 65) in a region.
Early evidence regarding COVID-19 suggests that the case rate mortality
is higher at older ages (e.g.~The Novel Coronavirus Pneumonia Emergency
Response Epidemiology Team, 2020). However, it is not clear that a
relatively large population of older adults necessarily translates into
higher transmission of the infection. The tool of choice in containing
the spread of the disease has been social distancing. In this respect,
the evidence from the field of transportation is that older adults tend
to travel less frequently, for shorter distances, and have higher rates
of immobility (e.g., Roorda et al., 2010; Morency et al., 2011; Sikder
and Pinjari, 2012). In other words, many older adults are, whether by
preference or otherwise, already in a form of social isolation. Social
distancing during the pandemic may actually reinforce that condition for
them, as suggested by the analysis of age-structured social contact in
India, China, and Italy of Sing and Adhikari (2020). Accordingly, our
expectation is that higher percentages of older adults will tend to have
similarly low levels of social contact (the age-structured matrix of
social contact in Spain is similar to Italy; see Prem et al., 2017).

Population density is also relevant since it directly affects the
contact patterns and contact rates between individuals in a population
(Hu et al., 2013). The evidence available suggests a positive
relationship between the transmission of COVID-19 and population density
(e.g.~cumulative incidence in urban areas like NYC). For this reason, we
anticipate a positive relationship between population density and the
incidence of the disease.

The last variable that we consider as a control is the presence of mass
transit systems in a province. Every province in Spain offers some form
of public transportation, however only five provinces have higher order
systems of mass mobility (e.g.~metro or subway), namely Barcelona,
Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been
hypothesized to relate to the spread of contagious disease by some
researchers using agent-based approaches and simulation (e.g., Perez and
Dragicevic, 2009; Wang et al., 2010), and while we find scant evidence
of a link in the literature, the idea is intuitively appealing. After
all, unlike the isolation that a car offers to travellers, most mass
transit system are cauldrons of social contact.

@@

The first reported case of COVID-19 in Spain was on January 31st, 2020,
when a German tourist in the Canary Islands tested positive for the
virus. However, it was still a few weeks before the first domestic case
was reported, on February 27th in Sevilla province (Andalusia). In a
short period of time, as testing started to ramp up, it became clear
that an outbreak was flaring. By March 11th the World Health
Organization (WHO) declared COVID-19 officially a pandemic. This
declaration marked a turning point for the public in Spain too. As of
March 13th, the number of cases of COVID-19 reported in Spain was 4,473,
with a majority of cases (1,990) concentrated in Madrid: these numbers
were at the time the worst outbreak in Europe after Italy. In response
to the situation, on March 13th the Spanish National Government declared
a state of emergency, to go into effect on Saturday March 14th. As part
of the state of emergency restrictions to most activities were imposed,
with the exception of essential services (e.g.~food, health) and some
economic subsectors of industry and construction. A few days later, on
March 17th, Spain closed its lands borders to allow entry only to
returnee nationals and permanent residents. The lockdown was further
hardened between March 30th and April 12th (including the Easter weekend
of April 10th-12th) and during this period only essential activities
were allowed. During this period, there was a dramatic reduction in
overall mobility, both within provinces as between
\footnote{https://www.mitma.gob.es/ministerio/covid-19/evolucion-movilidad-big-data/movilidad-provincial}.

@@

Our dataset includes information about the daily number of cases of
COVID-19 reported in Spain at the provincial level (NUTS3 in Eurostat
terminology) for the period between March 13th and April 11th,
inclusive. For our purposes, we consider positive cases reported, but
excluding symptomatic cases diagnosed by a doctor without a Polymerase
Chain Reaction (PCR) test. The Spanish National Government publishes
periodic updates at the regional level (NUTS2) and the information is
also released at the provincial level as part of a collaborative project
by geovoluntarios.com\footnote{https://www.geovoluntarios.org/},
ProvidencialData19\footnote{https://www.datoscovid.es/pages/providencialdata19},
and ESRI España. This information is compiled from several sources,
mainly the official web pages of the Spanish regional goverments, as
documented in Centro de Datos
Covid-19\footnote{https://www.datoscovid.es/pages/sobre-la-iniciativa}.
In addition, we consider two sets of explanatory variables. The first
one, and the focus of this research, is set of two environmental
variables, namely temperature and humidity, which are collected from
official sources (i.g., \emph{AEMET}, the state meteorology agency, and
\emph{MAPA}, the ministry of agriculture, fisheries, and food). The
second set provides some relevant controls for multivariate analysis,
and refers to economic and demographic attributes of the province (also
collected from official sources, i.e., INE, the national statistics
institute). Table \ref{tab:descriptive-statistics} shows the descriptive
statistics and the provenance of the data.

The spatial and temporal coverage of the data is as follows. Our dataset
begins on March 13, which is the first date when every province had
reported at least one case of COVID-19, and continues until April 11,
for a period of 30 days. The unit of analysis is the province. Provinces
are the equivalent of states, and are embedded in Autonomous
Communities. As an example, Cataluña is an Autonomous Community and
consists of four provinces, namely Barcelona, Gerona, Lerida, and
Tarragona. The size of the provinces is relatively large, as seen in
Table \ref{tab:descriptive-statistics}. The smallest province is
\(1,978.12km^2\) (this is smaller than Rhode Island in the U.S.) and the
largest province is \(21,767.93km^2\) (slightly smaller than New Jersey
in the U.S.). While this is a relatively large degree of spatial
aggregation, reporting on COVID-19 is not yet consistent at smaller
geographies, or cases are not reported at that level at all.

@@

\caption{\label{tab:descriptive-statistics}\label{tab:descriptive-statistics} Descriptive statistics}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l>{\raggedright\arraybackslash}p{15em}rrrr>{\raggedright\arraybackslash}p{5em}}
\toprule
Variable & Note & Min & Mean & Max & SD & Source\\
\midrule
\rowcolor{gray!6}  COVID-19 Incidence & Incidence in reported cases of SARS-19 per 100,000 people & 0.38 & 153.61 & 1149.36 & 186.23 & ProvidencialData19\\
Area & Area of province in sq.km & 1978.12 & 10118.79 & 21767.93 & 4.77 & INE\\
\rowcolor{gray!6}  GDPpc & GDP per capita in €1,000s & 16.67 & 22.51 & 36.00 & 4805.98 & INE\\
Older & Percentage of people aged 65 and older in the province & 15.16 & 21.03 & 31.36 & 3.95 & INE\\
\rowcolor{gray!6}  Population Density & Population density in the province in people per sq.km & 8.60 & 140.04 & 829.76 & 181.25 & INE\\
\addlinespace
Mean Temperature & Mean temperature in province by date, in °C & 1.00 & 12.18 & 23.20 & 3.67 & AEMET\\
\rowcolor{gray!6}  Humidity & Relative humidity in province by date & 2.00 & 77.82 & 100.00 & 10.37 & MAPA\\
Sunshine & Daily hours of sunshine in province by date & 0.00 & 5.74 & 12.40 & 3.96 & AEMET\\
\bottomrule
\multicolumn{7}{l}{\textit{Note: }}\\
\multicolumn{7}{l}{ProvidencialData19: https://www.datoscovid.es/pages/providencialdata19}\\
\multicolumn{7}{l}{INE (Instituto Nacional de Estadistica): https://www.ine.es/}\\
\multicolumn{7}{l}{AEMET (Agencia Estatal de Meteorologia): http://eportal.mapa.gob.es}\\
\multicolumn{7}{l}{MAPA (Ministerio de Agricultura, Pesca y Alimentacion): http://eportal.mapa.gob.es}\\
\end{tabular}}
\end{table}

@@

The Seemingly Unrelated Regression equations model (SUR hereafter) is a
multivariate econometric model used in different fields when the
structure of the data consists of cross-sections for different time
periods. The basis of this approach is well-known since the initial
works of Zellner (1962), and has become a popular methodology included
in several econometrics textbook (e.g., Greene, 2003). To our knowledge,
Anselin (1988) was the first author to discuss SUR from a spatial
perspective, in the context of spatio-temporal analysis. In his landmark
text, Anselin discussed a model made of ``an equation for each time
period, which is estimated for a cross section of spatial units''
(p.~141). From this milestone, a large body of research has developed to
extend the classical SUR into a spatial framework (e.g., Rey and
Montouri, 1999; Lauridsen et al., 2010; Le Gallo and Dall'Erba, 2006;
López et al., 2017).

@@

A key feature of the SUR model is the temporal dependence structure
among the vectors of residuals, namely:

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

Figure \ref{fig:descriptives-temperature} shows the distribution of the
environmental variables in Spain. For ease of visualization we aggregate
the provinces by Autonomous Community. Each box-and-whisker in the
figure represents the distribution of the variable for a community over
the 30-day period. In the plot, the communities have been sorted by
latitude, so that Principado de Asturias is the northernmost community,
and Canarias the southernmost. As seen in the figure, there is a
relatively wide range of values both within and between provinces over
the 30-day period examined. The top panel of the figure shows the
distribution of mean temperatures. The lowest mean temperature for a
community on any given day was 2.8°C, and the highest 22.4°C, for a
range of approximately 20 degrees. Likewise, there is a great deal of
variability in humidity, as seen in the middle panel of the figure,
where the lowest mean humidity for any community is 48.3 and the highest
is 99.6. Finally, the bottom panel displays mean daily hours of sunshine
in the community. This variable displays the most variability within
communities over time, but remains relatively constant across
communities. It is important to note that these are summaries by
community, and the actual values of these variables for the provinces
display somewhat more variability.

@@

It can be seen in the figure that temperature (in its three forms) has
the highest simple correlation with incidence of COVID-19. After
temperature, GDP per capita has the highest positive correlation with
the dependent variable. The distribution of these correlations is also
quite tight over the 30-day period of the study. Hours of sunshine tends
to have a moderately high correlation with incidence of COVID-19,
however the distribution of these correlations is more spread, and in
some cases strays into negative values. A similar thing happens with
humidity, which also tends to display more day to day variation in the
correlation with the dependent variable. The percentage of older adults
shows a relatively tight distribution of day-to-day correlations, and is
negative. Population density, in contrast, tends to be negative, but is
relatively spread, and on some days, the simple correlation between
density and incidence of COVID-19 is weakly positive. Overall for the
period under examination, the pairwise correlations between these
variables and incidence are significant at \(p<0.05\), with the
exception of the three \(Sunshine_hours\) variables.

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

Two variables had temporally constrained coefficients. The estimated
effect of GDP per capita is to increase the incidence of COVID-19 by
0.675\% for each percent increase of this variable (in €1,000s). In our
view, this is a measure of inertia, as provinces with higher GDP per
capita where among the first to see exponential growth in the pandemic.
Percentage of older adults has a negative effect, and each percent
increase in this variable is associated with a relatively small
reduction of the incidence of less than 1\%.
